### 2021-10-29 manuscript/hiv-gwas.tex
Update on Overleaf.

\documentclass[]{article}
\usepackage[margin=1in]{geometry}
\usepackage{biblatex}
\addbibresource{references.bib}
\usepackage{graphicx}
\usepackage{bm} % for bold math characters
\usepackage{booktabs} % for prettier hlines in table
\usepackage{subcaption} % for subfigures
\usepackage[affil-it]{authblk} % for footnote-style author affiliation
\usepackage{setspace} % for 1.5-spaced lines not including footnotes, floats
\usepackage{tikz} % for graphical model drawing
\usepackage{float} % to enforce figure positioning in subsection they're declared in
\usepackage{amsmath}
\usepackage{longtable} % for table that spans multiple pages
\usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions

@@

% How heritable is spVL (human side)
GWAS can be used to measure narrow-sense heritability from human genetic variants. Several GWAS have been done for spVL \parencite{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. The largest to-date by \citet{McLaren2015} measured the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%. All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}). Both associations are biologically relevant: the MHC encodes proteins that display viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry. In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that infected cells should be destroyed. On the other hand, CCR5 proteins are a signal to the virus that help it find the right cells to infect.

@@

Our premise is that spVL is the sum of several parts: direct host genetic effects, pathogen genetic effects, interaction effects between host and pathogen genetic factors, and environmental effects. Of these, only pathogen genetic effects should be heritable from one transmission partner to another \parencite{Mitov2018, Bertels2018}. To characterize these effects, we use a phylogenetic mixed model (PMM, \cite{Housworth2004}). PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts. In our case, the heritable part comes from pathogen genetic effects and the non-heritable part comes from the other effects. Then, viral genetic effects are modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny. The other, non-heritable effects are modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny (Figure \ref{fig:spVLModel}).

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.75\linewidth]{figures/model_figure}
		\caption{The POUMM statistical model for the spVL trait. (A) shows how the pathogen effect on spVL, $g_v$, evolves according to an OU process along (C) the pathogen phylogeny. The dashed line in (A) shows the expected value of the trait at each point in time under the model. (B) shows that the non-pathogen effect on spVL, $e$, is modeled as a Gaussian random variable. Each tip on the phylogeny corresponds to a sampled individual with an independent non-pathogen-effect drawn from the distribution shown in (B).}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

So far, spVL has been modelled using PMMs with two different types of random processes. The Brownian Motion (BM) process assumes that trait values are unbounded (i.e. spVL can attain any value). In contrast, the Ornstein-Uhlenbeck (OU) process assumes the trait fluctuates around an optimal value (i.e. extreme spVL values are very unlikely) \cite{Butler2004}. We argue the OU assumption is more realistic for spVL. \citet{Fraser2014} hypothesized that spVL is under stabilizing selection to maximize viral transmission. Accordingly, both \citet{Mitov2018} and \citet{Bertels2018} found higher statistical support for the OU process compared to BM for spVL evolution. Therefore, we assume the OU process. This model, called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM), is described in detail by \citet{Mitov2018}, but we review the main points here in the spVL context.

Under the POUMM, the spVL trait $z$ is the sum of direct host genetic effects $g_h$, viral genetic efects $g_v$, and environmental effects $\epsilon$. We can group the phylogeny-independent effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$: 
\begin{equation}
    z = g_{v} + e
\end{equation}
$g_v$ is a viral trait that evolves along the phylogeny according to an OU process. The OU process is defined by a stochastic differential equation with two terms. The first term represents a deterministic pull towards an optimal trait value and the second term represents stochastic fluctuations modelled by Brownian motion \cite{Butler2004}:
\begin{align}
\begin{split}
	&dg_v(t) = \alpha[\theta - g_v(t)]dt + \sigma dW_t \\
	&g_v(0) = g_0 
	\label{eq:OUprocess}
\end{split}
\end{align}

@@

$g_v$ evolves independently in descendent lineages after a divergence event in the phylogeny. The covariance between $g_v(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \parencite{Butler2004}: 
\begin{equation}
	Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big) = \frac{\sigma^2}{2\alpha}[e^{-\alpha d_{ij}}(1 - e^{-2\alpha t_{0(ij)}})]
	\label{eq:OUcov}
\end{equation}
Finally, we remember that $e$ is the non-heritable, environmental part of spVL. $e$ is modeled as a Gaussian random variable that is time- and phylogeny-independent. The expectation of $e$ is 0, meaning environmental effects are equally likely to raise or lower spVL from the virus-determined level. The parameter $\sigma_e^2$ measures the between-host variance of the environmental effect.
\begin{align}
\begin{split}
	E(e) = 0 \\
	Var(e) = \sigma^2_e
\end{split}
\end{align}

\subsection{Estimating and removing the virus-determined part of spVL}

Given the assumptions of the POUMM, we can derive a maximum-likelihood estimate for the virus-determined part of spVL for each individual in a cohort at the time they were sampled. Our goal is to estimate and remove this effect from measured spVL prior to GWAS.

@@

	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citet{Petersen2012}.
\end{proof}

Importantly, equation \ref{eq:MuG} is the maximum likelihood estimate for $\bm{g}_v$, the viral part of spVL, taking into account all available information - the viral phylogeny, the inferred POUMM parameters, and measured spVL values. This estimator is an inverse-variance weighted average of information from the POUMM evolutionary model ($\bm{\mu}_{OU}$) and measured spVL ($\bm{z}$). Intuitively, the estimate for $g_v$ is closer to measured spVL if spVL is not very heritable, and is closer to the expected value under the POUMM if spVL is highly heritable.

Given that we have an estimate for $\bm{g}_v$, we can finally estimate $\bm{e}$, representing the spVL value with viral effects removed: 
\begin{equation}
	\hat{\bm{e}} = \bm{z} - Exp(\bm{g}_v)
	\label{eq:EHat}
\end{equation}

\subsection{A new infectious disease GWAS framework}

We propose that by estimating and removing spVL variability due to the pathogen we can better measure the direct effects of human genetic variants on spVL. Our proposed framework is as follows:

\begin{enumerate}
	\item Sample host genotypes, host trait values, and pathogen genetic sequences from a cohort.
	\item Construct a pathogen phylogeny using the pathogen genetic sequences.
	\item Fit the POUMM to the trait values and the phylogeny. This produces estimates for parameters $g_0, \alpha, \theta, \sigma,\text{ and }\sigma_e$.
	\item Generate maximum-likelihood estimates for the pathogen-determined and environmental parts of the trait using equations \ref{eq:MuG} and \ref{eq:EHat}.
	\item Perform GWAS, using the estimated environmental part of the trait as the response variable instead of raw trait values. 
\end{enumerate}

@@

To test the theoretical best-case performance of our framework, we simulated data under the POUMM. Figure \ref{fig:sim-design} shows the simulation scheme and table \ref{tab:sim-params} gives the value or expression used for each parameter.

@@

First, we evaluated estimator accuracy with respect to $\bm{g}_h$, the sum of additive host genetic effects. This would be the ideal GWAS response variable to identify direct host determinants of spVL. We compute the accuracy of our proposed GWAS endpoint $\bm{\hat{e}}$ and the typical GWAS endpoint, measured spVL, using root mean squared error (RMSE). In our simulations we expect (scaled) measured spVL, $\bm{z} - \bar{\bm{z}}$, where $\bar{\bm{z}}$ is the average spVL, to be Gaussian distributed around $\bm{g}_h$ with variance equal to $\sigma^2_e + \sigma^2_{g_v} \approx 0.55$ (see table \ref{tab:sim-params} for simulation parameters). This should yield an RMSE of $\approx$ 0.74 for $\bm{z} - \bar{\bm{z}}$ as an estimate of $\bm{e}$, which matches simulation results (figure \ref{fig:simulationResults}A, scaled spVL). By incorporating information from the viral phylogenetic tree in calculating $\bm{\hat{e}}$, we were better able to estimate $\bm{e}$. The greatest improvement was possible in simulation scenarios in which spVL was highly heritable, meaning the virus contributed a great deal of variability to spVL. Estimates were also better in scenarios with lower selection strength since the correlation in $g_v$ among related tips is higher when $g_v$ can drift further from the long-term optimum and the stochastic fluctuations characterized by $\sigma$ are correspondingly lower to maintain the same viral genetic variance $\sigma^2_{g_v}$ and the same broad-sense heritability $H^2$.

@@

Next, we characterize the evolutionary scenarios in which our framework improves GWAS power. More accurate estimation of the non-viral part of spVL does not guarantee improved GWAS power. For example, if the viral part of spVL were constant across all viruses, association results (significance and effect size) would not be effected by a viral contribution to the trait. To determine under what evolutionary scenarios our method can improve GWAS power, we performed GWAS on the simulated data. Under our simulation set-up a standard GWAS would have $\approx$ 30\% power to detect a host variant explaining 1.25\% of the variability in spVL. This matches simulation results (figure \ref{fig:simulationResults}B, scaled spVL). Using our proposed framework, we were able to improve GWAS power in simulated scenarios with selection strength $<$ 10-20 time$^{-1}$ and spVL heritability $>$ 35-45\% (figure \ref{fig:simulationResults}B, $\bm{\hat{e}}$). If were were able to perfectly estimate $\bm{e}$, GWAS power would increase across all values of selection strength so long as the trait is highly heritable (figure \ref{fig:simulationResults}, $\bm{e}$). In summary, we show it is theoretically possible to improve GWAS power for heritable infectious disease traits by removing an estimated pathogen-determined effect.

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.75\linewidth]{figures/simulation_results}
		\caption{Results from the simulation study. We simulated pathogen genetic effects, host genetic effects, and environmental effects on the spVL trait under the POUMM with different trait heritability values (y-axis) and selection strengths (x-axis). For each simulated dataset, we applied our method to estimate the non-pathogen effects (host genetic \& environmental effects) and performed GWAS using these estimates. (A) shows how well we can estimate the host genetic effect using our method (left) compared to the whole trait value (scaled such that it has mean 0) (right). (B) shows the theoretical improvement to GWAS using the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) in different simulated scenarios. For comparison, we show GWAS results using the whole trait value (right). Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples for each heritability and selection strength value. The black point represents our estimates for the heritability and selection strength of spVL based on Swiss data. RMSE = Root mean square error, TPR = True positive rate.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

@@

Finally, we demonstrate our framework on empirical data from the Swiss HIV Cohort Study (SHCS). In the ``standard GWAS" we followed a typical GWAS methodology,  i.e. regressing measured spVL ($\bm{z}$) on variant dosage for each sampled host genetic variant. In the ``GWAS with phylogenetic correction", we use our proposed framework, i.e. regressing the estimated environmental part of spVL ($\bm{\hat{e}}$) on variant dosage. In agreement with previous GWAS for host genetic determinants of spVL, the top associations were in the MHC and \emph{CCR5} regions for both GWAS (figure \ref{fig:gwas-results}).

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures_archived/200207_gwas_results}
		\caption{Association results for standard GWAS and GWAS with a phylogenetic correction applied to the same data for HIV-1 set-point viral load. The pullout plots give regional association results from the GWAS with phylogenetic correction in the \emph{CCR5} region (positions 45.4 - 47Mb on chromosome 3) and the MHC region (positions 29.5 - 33.5Mb on chromosome 6). Base positions are with reference to genome build GRCh37. The color scale in the regional association plot is the difference in -log$_{10}$ p-value between the GWAS with phylogenetic correction and the standard GWAS. Red color means the phylogenetic correction decreased variant significance and blue color means the phylogenetic correction increased variant significance. The dashed lines show genome-wide significance (p-value = 5x10$^{-8}$).}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

@@

The top-associated SNPs from the \citet{McLaren2015} paper were excluded from the main analysis because of too-high missing rates (3\% for rs1015164 and 4\% for rs59440261). In order to directly compare our results with \citet{McLaren2015}, we ran an additional GWAS for these two SNPs anyways. Both SNPs became less significant with phylogenetic correction (table \ref{tab:comp-gwas}).

Again following the procedure of \citet{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region. Some HLA alleles also became more significant with a phylogenetic correction (table \ref{tab:lim-mclaren-comparison}).

\begin{table}[H]
	\begin{center}
		\begin{tabular}{p{1cm}p{2cm}p{2.5cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr} {llclcccc}
			\hline
			& & McLaren et al.  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
			\hline
			Region & Variant & p-value & Beta & p-value & Beta & p-value \\
			\hline
			MHC & rs59440261 & 2.0 x 10$^{-83}$ & -0.414 & 1.6x10$^{-11}$ & -0.229 & 1.2x10$^{-10}$ \\
			\emph{CCR5} & rs1015164 & 1.5 x 10$^{-19}$ & 0.148 & 1.3x10$^{-6}$ & 0.073 & 3.2x10$^{-5}$ \\			
			\hline
		\end{tabular}
		\caption{Top association results from \citet{McLaren2015} compared to association results in the current study.}
		\label{tab:comp-gwas}
	\end{center}
\end{table}

>>>>

\documentclass[]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{bm} % for bold math characters
\usepackage{booktabs} % for prettier hlines in table
\usepackage{subcaption} % for subfigures
\usepackage[affil-it]{authblk} % for footnote-style author affiliation
\usepackage{setspace} % for 1.5-spaced lines not including footnotes, floats
\usepackage{tikz} % for graphical model drawing
\usepackage{float} % to enforce figure positioning in subsection they're declared in
\usepackage{amsmath}
\usepackage{longtable} % for table that spans multiple pages
\usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions

\usepackage[backend=bibtex,style=nature]{biblatex} % Borrowing citation formatting from Nature
\bibliography{references}

@@

% How heritable is spVL (human side)
GWAS can be used to measure narrow-sense heritability from human genetic variants. Several GWAS have been done for spVL \parencite{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. The largest to-date by \cite{McLaren2015} measured the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%. All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}). Both associations are biologically relevant: the MHC encodes proteins that display viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry. In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that infected cells should be destroyed. On the other hand, CCR5 proteins are a signal to the virus that help it find the right cells to infect.

@@

Variation in spVL comes from several sources: direct host genetic effects, pathogen genetic effects, interaction effects between host and pathogen genetic factors, and other environmental effects. Of these, only pathogen genetic effects are heritable from one transmission partner to another \parencite{Leventhal2016}. To characterize these effects, we use a phylogenetic mixed model (PMM) \cite{Housworth2004}. PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts. In our case, pathogen genetic effects comprise the heritable part and all other effects comprise the non-heritable part. Then, the heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny. The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny (Figure \ref{fig:spVLModel}).

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.462\linewidth]{figures/model_figure}
		\caption{The POUMM statistical model for the spVL trait. (A) shows how the pathogen effect on spVL, $g_v$, evolves according to an OU process along (C) the pathogen phylogeny. The dashed line in (A) shows the expected value of the trait at each point in time under the model. (B) shows that the non-pathogen effect on spVL, $e$, is modeled as a Gaussian random variable. Each tip on the phylogeny corresponds to a sampled individual with an independent non-pathogen-effect drawn from the distribution shown in (B).}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

So far, PMMs with two types of random processes have been used to model spVL evolution. The Brownian Motion (BM) process assumes unbounded trait values, i.e. spVL can attain any value. The Ornstein-Uhlenbeck (OU) process assumes trait values fluctuate around an optimal value, i.e. extreme spVL values are unlikely. \cite{Mitov2018} and \cite{Bertels2018} previously showed the OU process has higher statistical support for spVL. This makes sense given that spVL is likely under stabilizing selection to maximize viral transmission potential \cite{Fraser2014}. Therefore, we assume the OU process. The full model is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) and is described in detail by \cite{Mitov2018}. Here, we review the main relevant points in the spVL context.

Under the POUMM, the spVL trait $z$ is the sum of viral genetic efects $g_v$, host genetic effects $g_h$, and other environmental or interaction effects $\epsilon$. We can group the non-heritable effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$: 
\begin{equation}
    z = g_{v} + e
\end{equation}
$g_v$ is a viral trait that evolves along the phylogeny according to an OU process. The OU process is defined by a stochastic differential equation with two terms. The first term represents a deterministic pull towards an optimal trait value and the second term represents stochastic fluctuations modelled by Brownian motion \cite{Butler2004}:
\begin{align}
\begin{split}
	&dg_v(t) = \alpha[\theta - g_v(t)]dt + \sigma dW_t \\
	&g_v(0) = g_0 
	\label{eq:OUprocess}
\end{split}
\end{align}

@@

$g_v$ evolves independently in descendent lineages after a divergence event in the phylogeny. The covariance between $g_v(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \parencite{Butler2004}: 
\begin{equation}
	Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big) = \frac{\sigma^2}{2\alpha}[e^{-\alpha d_{ij}}(1 - e^{-2\alpha t_{0(ij)}})]
	\label{eq:OUcov}
\end{equation}
Next, we remember that $e$ is the non-heritable, environmental part of spVL. $e$ is modeled as a Gaussian random variable that is time- and phylogeny-independent. The expectation of $e$ is 0, meaning environmental effects are equally likely to raise or lower spVL from the virus-determined level. The parameter $\sigma_e^2$ measures the between-host variance of the environmental effect.
\begin{align}
\begin{split}
	E(e) = 0 \\
	Var(e) = \sigma^2_e
\end{split}
\end{align}

Finally, broad-sense trait heritability can be calculated as the fraction of total trait variance that is heritable:
\begin{equation}
	H^2_{\bar{t}} = \frac{Var[g_v(t)]}{Var[g_v(t)] + Var(e)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_e}
\end{equation}

\subsection{Teasing apart pathogen and non-pathogen effects on spVL}

Given the assumptions of the POUMM, we can estimate a heritable pathogen effect on spVL and a non-heritable, host and environmental effect on spVL. Here, we derive a maximum-likelihood estimate for these values for individuals in a cohort, given measured spVL values and a pathogen phylogeny linking the infecting strains.

@@

	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \cite{Petersen2012}.
\end{proof}

Importantly, equation \ref{eq:MuG} is the maximum likelihood estimate for $\bm{g}_v$, the viral effect on spVL, taking into account all available information - measured spVL, the pathogen phylogeny, and inferred POUMM parameters. This estimator is an inverse-variance weighted average of measured spVL ($\bm{z}$) and information from the POUMM evolutionary model ($\bm{\mu}_{OU}$). In other words, $g_v$ will be closer to measured spVL if spVL is not very heritable. If spVl is highly heritable, $g_v$ will be closer to the expected value under the POUMM, i.e. take more information from the phylogenetic relationships between infecting strains.

Given the estimator we just derived for $\bm{g}_v$, we can now estimate $\bm{e}$, the spVL value $without$ pathogen effects: 
\begin{equation}
	\hat{\bm{e}} = \bm{z} - Exp(\bm{g}_v)
	\label{eq:EHat}
\end{equation}

We will use this value to try to improve upon standard GWAS methods in infectious disease.

\subsection{A POUMM-based GWAS framework for infectious disease}

We propose to improve standard GWAS for infectious diseases by estimating and removing trait variability due to pathogen genetic effects. Our new framework is as follows:

\begin{enumerate}
	\item Sample host genotypes, trait values, and pathogen genome sequence data from a cohort.
	\item Construct a pathogen phylogeny using the pathogen genome sequences.
	\item Estimate the parameters of the POUMM based on the trait values and the pathogen phylogeny. This can be done with e.g. the R package POUMM \cite{Mitov2017a-POUMM}.
	\item Generate maximum-likelihood estimates for the pathogen and corresponding non-pathogen effects on the trait using equations \ref{eq:MuG} and \ref{eq:EHat}.
	\item Perform GWAS with only the non-pathogen effects on the trait as the response variable.
\end{enumerate}

@@

To test the theoretical best-case performance of our method, we simulated data under the POUMM and applied our framework to the simulated data. Figure \ref{fig:sim-design} shows the simulation scheme and table \ref{tab:sim-params} gives the value or expression for each parameter. In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength. To maintain the same heritability while varying selection strength, we counter-balanced by varying the intensity of stochastic evolutionary fluctuations accordingly. We fixed other variables to plausible values based on the spVL literature.

@@

First, we evaluated how well our method estimated the additive host genetic effects from the simulated data. Additive host genetic effects represent an ideal baseline for GWAS that seek to identify host genetic determinants of a trait. Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can generate more accurate estimates than the trait value alone. To be fair, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor. In the supplemental material, we show why the scaled trait value is expected to have an RMSE of approximately 0.74 under our simulation set-up. By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable and under low selection pressure. This is because high heritability means the pathogen greatly effects the trait value. Then, when the trait is under weak selection, pathogen effects can drift far from the long-term optimum and stochastic fluctuations are low to maintain the same heritability. Thus, our method performs well when an infectious disease trait is highly heritable and trait values are highly correlated amongst closely related members of a transmission chain, but vary widely within the entire cohort.

@@

Next, we characterized the evolutionary scenarios under which our framework can actually improve GWAS power. We used the true positive rate (TPR) to evaluate the fraction of simulated host genetic variants with direct effects we could recover as being significantly associated with the trait. We performed three different GWAS for each simulated dataset: the first represents an ideal in which we can exactly know and remove pathogen effects from trait values, the second is using our method to estimate this value and remove it, and the third represents a standard GWAS using the unmodified trait value. Figure \ref{fig:simulationResults}B shows that our framework can improve the TPR in simulated scenarios where selection strength $<$ 10 time$^{-1}$ and heritability $>$ 45\%. If were were able to perfectly estimate and remove pathogen effects on a trait, the TPR would increase across all values of selection strength so long as the trait is more than marginally heritable. We estimate approximately 25\% to be the heritability threshold above which GWAS power is negatively impacted by pathogen effects. In summary, we show it is theoretically possible to improve GWAS power for heritable infectious disease traits by estimating and removing pathogen effects using information from the pathogen phylogeny.

% More accurate estimation of the non-viral part of spVL does not guarantee improved GWAS power. For example, if the viral part of spVL were constant across all viruses, association results (significance and effect size) would not be effected by a viral contribution to the trait. 
% Under our simulation set-up a standard GWAS would have $\approx$ 30\% power to detect a host variant explaining 1.25\% of the variability in spVL. This matches simulation results (figure \ref{fig:simulationResults}B, scaled spVL).

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.75\linewidth]{figures/simulation_results}
		\caption{Results from the simulation study. We simulated host, pathogen, and environmental effects on a trait under the POUMM with different heritability (y-axis) and selection strength (x-axis) parameters. For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values. (A) shows that our method (left) can generate more accurate estimates of host genetic effects than the trait value, scaled by its mean (right). (B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the unmodified trait value (right). Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples. The black point represents our estimates for the heritability and selection strength of spVL based on Swiss data. RMSE = Root mean square error, TPR = True positive rate.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

@@

Finally, we applied our framework to empirical data from the Swiss HIV Cohort Study (SHCS). In the ``standard GWAS" we followed a typical GWAS methodology,  i.e. regressing measured spVL ($\bm{z}$) on variant dosage for each sampled host genetic variant. In the ``GWAS with phylogenetic correction", we use our proposed framework, i.e. regressing the estimated environmental part of spVL ($\bm{\hat{e}}$) on variant dosage. In agreement with previous GWAS for host genetic determinants of spVL, the top associations were in the MHC and \emph{CCR5} regions for both GWAS (figure \ref{fig:gwas-results}).

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.7\linewidth]{figures/gwas_results}
		\caption{Association results for standard GWAS and GWAS with a phylogenetic correction applied to the same data for HIV-1 set-point viral load. The pullout plots give regional association results from the GWAS with phylogenetic correction in the \emph{CCR5} region (positions 45.4 - 47Mb on chromosome 3) and the MHC region (positions 29.5 - 33.5Mb on chromosome 6). Base positions are with reference to genome build GRCh37. The color scale in the regional association plot is the difference in -log$_{10}$ p-value between the GWAS with phylogenetic correction and the standard GWAS. Red color means the phylogenetic correction decreased variant significance and blue color means the phylogenetic correction increased variant significance. The dashed lines show genome-wide significance (p-value = 5x10$^{-8}$).}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

@@

The top-associated SNPs from the \cite{McLaren2015} paper were excluded from the main analysis because of too-high missing rates (3\% for rs1015164 and 4\% for rs59440261). In order to directly compare our results with \cite{McLaren2015}, we ran an additional GWAS for these two SNPs anyways. Both SNPs became less significant with phylogenetic correction (table \ref{tab:comp-gwas}).

Again following the procedure of \cite{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region. Some HLA alleles also became more significant with a phylogenetic correction (table \ref{tab:lim-mclaren-comparison}).

\begin{table}[H]
	\begin{center}
		\begin{tabular}{p{1cm}p{2cm}p{2.5cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr} {llclcccc}
			\hline
			& & McLaren et al.  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
			\hline
			Region & Variant & p-value & Beta & p-value & Beta & p-value \\
			\hline
			MHC & rs59440261 & 2.0 x 10$^{-83}$ & -0.414 & 1.6x10$^{-11}$ & -0.229 & 1.2x10$^{-10}$ \\
			\emph{CCR5} & rs1015164 & 1.5 x 10$^{-19}$ & 0.148 & 1.3x10$^{-6}$ & 0.073 & 3.2x10$^{-5}$ \\			
			\hline
		\end{tabular}
		\caption{Top association results from \cite{McLaren2015} compared to association results in the current study.}
		\label{tab:comp-gwas}
	\end{center}
\end{table}

\begin{table}[H]
	\begin{center}
	\begin{tabular} {p{3cm}p{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}
		\hline
		 & \multicolumn{2}{l}{McLaren et al.}  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
		 \hline
		Allele & Beta & p-value & Beta & p-value & Beta & p-value \\
		\hline	
		HLA-B*57:01 & -0.840 & $2.69 \times 10^{-86}$ & -0.479 & $1.11 \times 10^{-12}$ & -0.268 & $5.43 \times 10^{-12}$ \\ 
		\rowcolor{Gray}
		HLA-C*06:02 & -0.436 & $7.32 \times 10^{-50}$ & -0.173 & $1.10 \times 10^{-4}$ & -0.107 & $3.15 \times 10^{-5}$ \\ 
		HLA-B*27:05 & -0.417 & $4.32 \times 10^{-20}$ & -0.244 & $9.59 \times 10^{-4}$ & -0.122 & $4.40 \times 10^{-3}$ \\ 
		\rowcolor{Gray}
		HLA-C*07:01 & 0.195 & $1.86 \times 10^{-14}$ & 0.129 & $7.55 \times 10^{-3}$ & 0.086 & $1.99 \times 10^{-3}$ \\ 
		HLA-C*07:02 & 0.194 & $3.75 \times 10^{-13}$ & 0.095 & $3.02 \times 10^{-2}$ & 0.052 & $3.98 \times 10^{-2}$ \\ 
		HLA-B*07:02 & 0.200 & $5.17 \times 10^{-13}$ & 0.129 & $5.29 \times 10^{-3}$ & 0.072 & $7.06 \times 10^{-3}$ \\ 
		\rowcolor{Gray}
		HLA-B*08:01 & 0.206 & $2.94 \times 10^{-11}$ & -0.009 & $9.20 \times 10^{-1}$ & 0.015 & $7.75 \times 10^{-1}$ \\ 
		\rowcolor{Gray}
		HLA-C*08:02 & -0.307 & $8.76 \times 10^{-11}$ & 0.324 & $8.55 \times 10^{-1}$ & 0.400 & $6.95 \times 10^{-1}$ \\ 
		HLA-C*04:01 & 0.183 & $9.07 \times 10^{-11}$ & 0.124 & $3.04 \times 10^{-3}$ & 0.058 & $1.67 \times 10^{-2}$ \\ 
		\rowcolor{Gray}
		HLA-B*13:02 & -0.346 & $3.70 \times 10^{-10}$ & -0.188 & $1.03 \times 10^{-2}$ & -0.112 & $7.90 \times 10^{-3}$ \\ 
		\rowcolor{Gray}
		HLA-B*14:02 & -0.318 & $2.10 \times 10^{-9}$ & 0.067 & $8.12 \times 10^{-1}$ & 0.118 & $4.71 \times 10^{-1}$ \\ 
		\rowcolor{Gray}
		HLA-C*12:02 & -0.478 & $4.02 \times 10^{-8}$ & 0.111 & $7.63 \times 10^{-1}$ & 0.094 & $6.58 \times 10^{-1}$ \\
		\hline
	\end{tabular}
	\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \cite{McLaren2015} compared to results from the current study. The table is truncated to show only alleles that reached genome-wide significance in the McLaren et al. study. The full table is available in the supplement (table \ref{tab:full-mclaren-comparison}). Grey rows are alleles that increased in significance with a phylogenetic correction.}
	\label{tab:lim-mclaren-comparison}
	\end{center}
\end{table}
